The polynomial representation of the type $A_{n - 1}$ rational Cherednik algebra in characteristic $p \mid n$

### 2015-05-19 formalwriteup.tex
Line edits, Dunkl

The rational Cherednik algebra $\HH_{\hbar, c}(\h)$ admits a $\ZZ_{\geq 0}$-graded representation on the polynomial ring $A = \Sym(\h^*)$, known as the \textit{polynomial representation}. In particular, for any $f \in A$ we can take a partial derivative $\frac{\partial f}{\partial x_i}$ by noting that $A=k[x_0,\dots,x_{n-2}]$ with $x_{n-1}=-x_0-\dots-x_{n-2}$; then we can write $f$ as a polynomial and take its partial derivative by any $x_i$ with $0 \le i \le n-2$.

The actions of $\Sym(\h^*)$ and $k[S_n]$ on $A$ are by left multiplication and the $S_n$ action on $\h^*$, respectively.  The action of $\Sym(\h)$ is implemented by letting $y \in \h$ act via the \textit{Dunkl operator} $D_y$ defined by
\begin{align*}
D_y(f) = \hbar \partial_y f  - \sum_{k=0}^{n-1} \sum_{l=0}^{n-1} c  \langle y, x_k-x_l \rangle \frac{(1-s_{kl}) f}{x_k - x_l},
\end{align*}
where we note that $\frac{(1-s_{kl}) f}{x_k - x_l}$ is a polynomial for $f \in A$. Explicitly, for 
\[
D_i = \hbar\frac{\partial f}{\partial x_i} -c \sum_{k \ne i} \frac{(1-s_{ki})f}{x_i-x_k}
\]
we have $D_{y_i - y_j} = D_i - D_j$.

\todo[inline]{Fill this in, $D_{y_i}$ is not well-defined because $y_i$ is not in $\h$}

@@

Since $x_{n-1}=-x_0-\dots-x_{n-2}$, we see that
\[
\frac{\partial g}{\partial x_i}=-\frac{zg}{1-x_iz}+\frac{zg}{1-x_{n-1}z} \text{ for all $0 \le i < n-1$}.
\]
Then we see that for all $0 \le i < n-1$ we have 
\begin{align*}
\frac{\partial F}{\partial x_i}%&=\frac{\partial}{\partial x_i}\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)\\
&=\left(\sum_{k=1}^{p-1}k\binom{c}{k}(g-1)^{k-1}\right)\frac{\partial g}{\partial x_i}\\
&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=1}^{p-1}k\binom{c}{k}(g-1)^{k-1}\right)g\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=0}^{p-2}(k+1)\binom{c}{k+1}(g-1)^{k}\right)g\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)(g-1+1)\\
&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)+\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k+1}\right)\right)\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)+\left(\sum_{k=1}^{p-1}c\binom{c-1}{k-1}(g-1)^{k}\right)\right)\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(-c\binom{c-1}{p-1}(g-1)^{p-1}+\sum_{k=0}^{p-1}c\binom{c}{k}(g-1)^{k}\right)\\
&=\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)\left(-\binom{c-1}{p-1}(g-1)^{p-1}+F(z)\right).
\end{align*}
We define
\[
G_i(z)=\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)\left(-\binom{c-1}{p-1}(g-1)^{p-1}\right).
\]
Then we see that
\[
\frac{\partial F}{\partial x_i}=G_i(z)+\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)F(z).
\]

Then it remains only to show that $[z^l]G_i(z)=0$ for $l=0,\dots,p$. To show this, we need only show that $G_i(z)=z^{p+1}g_i(z)$ for some power series $v(z)$.

We consider the expansion of $g(z)$; this is $g(z)=\prod_{j=0}^{n-1} (1-x_jz) = 1-z\left(\sum_i x_i\right)+z^2e_2(x_i)+\dots+z^ne_n(x_i)$ where $e_2,\dots,e_n$ are polynomials in the $x_i$. In particular we note that $x_{n-1}=-x_0-\dots-x_{n-2}$ by hypothesis, so $z^2 \mid g(z)-1$.

Then by the definition of $G_i(z)$, we see that $z(g-1)^{p-1} \mid G_i(z)$, so $z^{2p-1} \mid V(z)$. Since $p \ge 2$, it is clear that $z^{p+1} \mid G_i(z)$, so we can write $G_i(z)=z^{p+1}g_i(z)$ for some power series $g_i(z)$ as desired.

\end{proof}

@@

\begin{lemma} \label{lem:resbyparts}
For any $y_0,\dots,y_{r-1}$ distinct elements of $k$ such that $\prod_i y_i \ne 0$ and $m_i$ positive integers such that $\sum_i m_iy_i=0$, define $g(z), F(z)$ as in Lemma~\ref{lem:dFdz2}. Let 
\[
a(z)=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z).
\].
Then we have
\begin{enumerate}
\item $a'(z)=z^pb(z)-\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)$ for some power series $b(z)$;
\item For any integer $l \le p-r$, $[z^l]a(z)$ is a linear combination of $[z^{l+1}]a(z),\dots,[z^{l+r}]a(z)$.\end{enumerate}

\end{lemma}

\begin{proof} We begin by computing $a'(z)$. By Lemma~\ref{lem:dFdz2} we have
\[
F'(z)=V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right),
\]
where $[z^l]V(z)=0$ for $0 \le l \le p-1$. 
We now compute
\begin{align*}
a'(z)&=\frac{d}{dz}\left(\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z)\right)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F'(z)+\sum_{j=0}^{r-1}\frac{y_j}{(1-y_jz)\prod_{i=0}^{r-1}(1-y_iz)}F(z)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}\left(V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}a(z)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}-\left(\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)\right).
\end{align*}
Since $[z^l]V(z)=0$ for $0 \le l \le p-1$ we see that $\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}=z^pb(z)$ for some power series $b(z)$. Therefore
\[
a'(z)=z^pb(z)-\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)
\]
for some power series $b(z)$.

For any integer $l$, define the Laurent polynomial 
\[
h_l(z)=z^{-l-r}\prod_i(1-y_iz).
\]
Notice that
\begin{align*}
\frac{d}{dz}( h_l(z) a(z)) &= -(l + r)z^{-1} h_l(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_l(z) a(z) + h_l(z) a'(z) \\
&=  -(l + r)z^{-1} h_l(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_l(z) a(z) - h_l(z) \sum_j \frac{(m_j c - 1)y_i}{1 - y_ijz} a(z)  + z^ph_l(z) b(z)\\
&= -(l + r)z^{-1} h_l(z) a(z) - \sum_i \frac{y_im_i c}{1 - y_iz} h_l(z) a(z) + z^ph_l(z) b(z)\\
&= -\left((l + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_l(z) a(z) + z^ph_l(z) b(z).
\end{align*}
We note that 
\[
-\left((l + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_l(z)
\]
is a Laurent polynomial with lowest degree term 
\[
- z^{-l-r-1} \left(l + r + \sum_i y_i m_i c\right) = - z^{-l-r-1}(l+ r)
\]
and highest degree term 
\[
- z^{-l - 1} (-1)^r  \prod_i y_i\left((l + r) - \sum_i m_i c\right).
\]
The lowest degree term of the formal power series $z^ph_l(z)b(z)$ has degree at least $p-l-r$. We note that we must have $[z^{-1}]\left(\frac{d}{dz}( h_l(z) a(z))\right)=0$, so for $p-l-r \ge 0$, we see that $[z^{-1}]\left(\frac{d}{dz}( h_l(z) a(z))\right)=0$ is a linear combination of $[z^l]a(z),[z^{l+1}]a(z),\dots,[z^{l+r}]a(z)$ in which $[z^l]a(z)$ has nonzero coefficient $ (-1)^r  \prod_i y_i\left((l + r) - \sum_i m_i c\right)$. The latter coefficient is nonzero since $c$ is generic and $\prod_i y_i \ne 0$. Therefore $[z^l]a(z)$ is a linear combination of $[z^{l+1}]a(z),\dots,[z^{l+r}]a(z)$ when $l \le p-r$.

\end{proof}

>>>>

The rational Cherednik algebra $\HH_{\hbar, c}(\h)$ admits a $\ZZ_{\geq 0}$-graded representation on the polynomial ring $A = \Sym(\h^*)$, known as the \textit{polynomial representation}.  The actions of $\Sym(\h^*)$ and $k[S_n]$ on $A$ are by left multiplication and the $S_n$ action on $\h^*$, respectively.  The action of $\Sym(\h)$ is implemented by letting $y \in \h$ act via the \textit{Dunkl operator} $D_y$ defined by
\begin{align*}
D_y(f) = \hbar \partial_y f  - \sum_{k=0}^{n-1} \sum_{l=0}^{n-1} c  \langle y, x_k-x_l \rangle \frac{(1-s_{kl}) f}{x_k - x_l},
\end{align*}
where we note that $\frac{(1-s_{kl}) f}{x_k - x_l}$ is a polynomial for $f \in A$. Explicitly, we have  
\[
D_{y_i - y_j} = 
\]

\todo[inline]{Remove $D_i$ and replace the quantity above}

@@

Since $x_{n-1}=-x_0-\dots-x_{n-2}$, we see that
\[
\frac{\partial g}{\partial x_i}=-\frac{zg}{1-x_iz}+\frac{zg}{1-x_{n-1}z} \text{ for all $0 \le i < n-1$}.
\]
Then we see that for all $0 \le i < n-1$ we have 
\begin{align*}
\frac{\partial F}{\partial x_i}%&=\frac{\partial}{\partial x_i}\left(\sum_{k=0}^{p-1} \binom{c}{k} (g-1)^k\right)\\
&=\left(\sum_{k=1}^{p-1}k\binom{c}{k}(g-1)^{k-1}\right)\frac{\partial g}{\partial x_i}\\
&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=1}^{p-1}k\binom{c}{k}(g-1)^{k-1}\right)g\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=0}^{p-2}(k+1)\binom{c}{k+1}(g-1)^{k}\right)g\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)(g-1+1)\\
&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}+\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k+1}\right)\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(\left(\sum_{k=0}^{p-2}c\binom{c-1}{k}(g-1)^{k}\right)+\left(\sum_{k=1}^{p-1}c\binom{c-1}{k-1}(g-1)^{k}\right)\right)\\
%&=\left(-\frac{z}{1-x_iz}+\frac{z}{1-x_{n-1}z}\right)\left(-c\binom{c-1}{p-1}(g-1)^{p-1}+\sum_{k=0}^{p-1}c\binom{c}{k}(g-1)^{k}\right)\\
&=\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)\left(-\binom{c-1}{p-1}(g-1)^{p-1}+F(z)\right).
\end{align*}
We define
\[
G_i(z)=\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)\left(-\binom{c-1}{p-1}(g-1)^{p-1}\right).
\]
Then we see that
\[
\frac{\partial F}{\partial x_i}=G_i(z)+\left(-\frac{zc}{1-x_iz}+\frac{zc}{1-x_{n-1}z}\right)F(z).
\]
It remains only to show that $[z^l]G_i(z)=0$ for $l=0,\dots,p$. To show this, we need only show that $G_i(z)=z^{p+1}g_i(z)$ for some power series $v(z)$.  Consider the expansion 
\[
g(z)=\prod_{j=0}^{n-1} (1-x_jz) = 1-z\left(\sum_i x_i\right)+z^2e_2(x_0, \ldots, x_{n-1})+\dots+z^ne_n(x_0, \ldots, x_{n - 1}).
\]
Recall that $\sum_i x_i = 0$, so $z^2 \mid g(z)-1$.  By definition of $G_i(z)$, we see that $z(g-1)^{p-1} \mid G_i(z)$, so $z^{2p-1} \mid V(z)$. Since $p \ge 2$, this implies $z^{p+1} \mid G_i(z)$, giving the desired.
\end{proof}

@@

\begin{lemma} \label{lem:resbyparts}
For any $y_0,\dots,y_{r-1}$ distinct elements of $k$ such that $\prod_i y_i \ne 0$ and $m_i$ positive integers such that $\sum_i m_iy_i=0$, define $g(z), F(z)$ as in Lemma~\ref{lem:dFdz2}. Let 
\[
a(z)=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z).
\].
Then we have
\begin{enumerate}
\item $a'(z)=z^pb(z)-\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)$ for some power series $b(z)$;
\item For any integer $l \le p-r$, $[z^l]a(z)$ is a linear combination of $[z^{l+1}]a(z),\dots,[z^{l+r}]a(z)$.
\end{enumerate}
\end{lemma}
\begin{proof} We begin by computing $a'(z)$. By Lemma~\ref{lem:dFdz2} we have
\[
F'(z)=V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right),
\]
where $[z^l]V(z)=0$ for $0 \le l \le p-1$. 
We now compute
\begin{align*}
a'(z)&=\frac{d}{dz}\left(\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F(z)\right)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}F'(z)+\sum_{j=0}^{r-1}\frac{y_j}{(1-y_jz)\prod_{i=0}^{r-1}(1-y_iz)}F(z)\\
&=\frac{1}{\prod_{i=0}^{r-1}(1-y_iz)}\left(V(z)+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}F(z)\right)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}+\left(\sum_j \frac{-cm_jy_j}{1-y_jz}a(z)\right)+\sum_{j=0}^{r-1}\frac{y_j}{1-y_jz}a(z)\\
&=\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}-\left(\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)\right).
\end{align*}
Since $[z^l]V(z)=0$ for $0 \le l \le p-1$ we see that $\frac{V(z)}{\prod_{i=0}^{r-1}(1-y_iz)}=z^pb(z)$ for some power series $b(z)$. Therefore
\[
a'(z)=z^pb(z)-\sum_j \frac{(m_jc-1)y_j}{1-y_jz}a(z)
\]
for some power series $b(z)$.

For any integer $l$, define the Laurent polynomial 
\[
h_l(z)=z^{-l-r}\prod_i(1-y_iz).
\]
Notice that
\begin{align*}
\frac{d}{dz}( h_l(z) a(z)) &= -(l + r)z^{-1} h_l(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_l(z) a(z) + h_l(z) a'(z) \\
&=  -(l + r)z^{-1} h_l(z) a(z) - \sum_i \frac{y_i}{1 - y_iz} h_l(z) a(z) - h_l(z) \sum_j \frac{(m_j c - 1)y_i}{1 - y_ijz} a(z)  + z^ph_l(z) b(z)\\
&= -(l + r)z^{-1} h_l(z) a(z) - \sum_i \frac{y_im_i c}{1 - y_iz} h_l(z) a(z) + z^ph_l(z) b(z)\\
&= -\left((l + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_l(z) a(z) + z^ph_l(z) b(z).
\end{align*}
We note that 
\[
-\left((l + r) z^{-1} + \sum_i \frac{y_im_i c}{1 - y_iz}\right) h_l(z)
\]
is a Laurent polynomial with lowest degree term 
\[
- z^{-l-r-1} \left(l + r + \sum_i y_i m_i c\right) = - z^{-l-r-1}(l+ r)
\]
and highest degree term 
\[
- z^{-l - 1} (-1)^r  \prod_i y_i\left((l + r) - \sum_i m_i c\right).
\]
The lowest degree term of the formal power series $z^ph_l(z)b(z)$ has degree at least $p-l-r$. We note that we must have $[z^{-1}]\left(\frac{d}{dz}( h_l(z) a(z))\right)=0$, so for $p-l-r \ge 0$, we see that $[z^{-1}]\left(\frac{d}{dz}( h_l(z) a(z))\right)=0$ is a linear combination of $[z^l]a(z),[z^{l+1}]a(z),\dots,[z^{l+r}]a(z)$ in which $[z^l]a(z)$ has nonzero coefficient $ (-1)^r  \prod_i y_i\left((l + r) - \sum_i m_i c\right)$. The latter coefficient is nonzero since $c$ is generic and $\prod_i y_i \ne 0$. Therefore $[z^l]a(z)$ is a linear combination of $[z^{l+1}]a(z),\dots,[z^{l+r}]a(z)$ when $l \le p-r$. 
\end{proof}
